\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-03 papers/jfp/8-engineering.tex
Rework polymorphic values

\subsection{Polymorphic values}\label{sec-polymorphism}

Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
system that works directly on \hs{k} and \hs{v} values. However, some build
systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
values, allowing types that are stored only in the \Shake's build information.
\Shake users have remarked that polymorphism provides a much easier expression
of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.

As one example of richer key/value types, consider the version of
\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
to define an \emph{oracle} rule as per Mitchell~\shortcite{mitchell2012shake}
whose value is the result of running \cmd{gcc -}\cmd{-version} and which is
volatile, making the \cmd{gcc} version something that can be depended upon. Of
course, provided the build can express volatile dependencies and supports
cutoff, the version number could equally be written to a file and used in a
similar way, albeit at the cost of introducing additional correctness
invariants, e.g. that the file \cmd{gcc-version.txt} does indeed contains two
dot-separated integers.

Generalised algebraic data types, or GADTs~\cite{spj2006gadts}, can be used to
extend our models to support ``typed'' build tasks. The idea is to replace
the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
more polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The idea is best explained by way of an example:

@@

data Key a where
    File       :: FilePath -> Key String
    GccVersion :: Key Version
\end{minted}
\vspace{1mm}

\noindent
Here we extend the usual mapping from file paths to file contents with an
additional key type \hs{GccVersion} mapped to a \hs{Version}. The task
abstraction needs to be adjusted to cope with such keys (the suffix \hs{T}
stands for ``typed''):

@@

\noindent
The changes compared to the definition in~\S\ref{sec-task} are minimal: (i) the
\hs{TaskT} now uses a typed \hs{Fetch} callback (we define a separate type
synonym only for readability), and (ii) the type of \hs{TasksT} is now
polymorphic in \hs{v} instead of being parameterised by a concrete \hs{v}. The
example below demonstrates how \hs{fetch} can be used to retrieve dependencies
of different types: the rule \cmd{release.txt} concatenates the contents of two
\hs{File}s, while the rule \cmd{main.o} uses the numeric \hs{GccVersion} to
determine the way the \hs{source} file should be compiled.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
example :: TasksT Monad Key
example (File "release.txt") = Just $ TaskT $ \fetch -> do
    readme  <- fetch (File "README")
    license <- fetch (File "LICENSE")
    return (readme ++ license)
example (File "main.o") = Just $ TaskT $ \fetch -> do
    let source = "main.c"
    version <- fetch GccVersion
    if version >= Version 8 0 then compileNew source
                              else compileOld source
example _ = Nothing
\end{minted}
\vspace{1mm}

>>>>

\subsection{Key-dependent value types}\label{sec-polymorphism}

Key-dependent value types allows a single build system to work with multiple different
types of values, where the type of any particular value is determined by the key. As an example
of why this might be useful, consider a build system that can depend on files
(whose contents are strings) or system executables (represented by their version number) --
using a single type for both values reduces type safety.
\Shake permits such key-dependent value types, for example the \emph{oracle} rule from
Mitchell~\shortcite{mitchell2012shake}, and users have remarked that this additional type safety
provides a much easier expression of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.

We can encode key-dependent value types using generalised algebraic data types,
or GADTs~\cite{spj2006gadts}. The idea is to replace
the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
more polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The variable \hs{k} has changed from kind \hs{*} (a type),
to \hs{* -> *} (a type function), permitting the key to constrain the type of the value.
The idea is best explained by way of an example:

@@

data Key a where
    File    :: FilePath -> Key String
    Program :: String -> Key Version
\end{minted}
\vspace{1mm}

\noindent
Here we extend the usual mapping from file paths to file contents with an
additional key type \hs{Program} which maps the name of a program to its installed \hs{Version}. The task
abstraction needs to be adjusted to cope with such keys (the suffix \hs{T}
stands for ``typed''):

@@

\noindent
The changes compared to the definition in~\S\ref{sec-task} are minimal: (i) the
\hs{TaskT} now uses a typed \hs{Fetch} callback (we define a separate type
synonym only for readability), and (ii) the type of \hs{TasksT} is now
polymorphic in \hs{v} instead of being parameterised by a concrete \hs{v}. The
example below demonstrates how \hs{fetch} can be used to retrieve dependencies
of different types: the rule \cmd{release.txt} concatenates the contents of two
\hs{File}s, while the rule \cmd{main.o} uses the numeric \hs{Program "gcc"} to
determine how the \hs{source} file should be compiled.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
example :: TasksT Monad Key
example (File "release.txt") = Just $ TaskT $ \fetch -> do
    readme  <- fetch (File "README")
    license <- fetch (File "LICENSE")
    return (readme ++ license)
example (File "main.o") = Just $ TaskT $ \fetch -> do
    let source = "main.c"
    version <- fetch (Program "gcc")
    if version >= Version 8 0 then compileNew source
                              else compileOld source
example _ = Nothing
\end{minted}
\vspace{1mm}

Note that like all key-dependent tasks, this example could be expressed
without key-dependence, at the cost of some type safety. As we'll see in
\S\ref{sec-multiple-outputs} and \S\ref{sec-tracking-aspects}, using key-dependent
value types can make it easier to work with more complicated tasks.

### 2019-05-03 papers/jfp/9-related.tex
Rework polymorphic values

\item \Ninja~\cite{ninja} combines the \hs{topological} scheduler of \Make with
the verifying traces of \Shake~--~our associated implementation provides such a
combination. \Ninja~is also capable of modelling build rules that produce
multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.

@@

\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
matches \Shake at the level of detail given here, differing only on aspects like
polymorphic dependencies~\S\ref{sec-polymorphism}.

>>>>

\item \Ninja~\cite{ninja} combines the \hs{topological} scheduler of \Make with
the verifying traces of \Shake~--~our associated implementation provides such a
combination. \Ninja~is also capable of modelling build rules that produce
multiple results, a limited form of multiple value types \S\ref{sec-polymorphism}.

@@

\item \Redo~\cite{redo-idea,grosskurth2007redo,redo} almost exactly
matches \Shake at the level of detail given here, differing only on aspects like
rules producing multiple files~\S\ref{sec-polymorphism}.
